### 2018-07-10 latex/discussion.tex
Incorporated changes suggested by Dorothée

Our findings shed also new light on the potential physiological role of NNMT, which has earlier been recognized as potential marker for some types of cancer (e.g. \cite{Okamura1998}). It might well be that increased NNMT expression in cancer might serve to remove Nam produced by increased NAD-dependent signalling. To maintain high NAD concentrations a simultaneous overexpression of NamPRT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}. As NNMT does, however, provide an advantage as long as NamPRT affinity is sufficiently high, this might provide an useful avenue for treatment, especially for types of cancer that express NNMT at a high level. These would be more susceptible to competitive inhibitors of NamPRT, and several of those inhibitors are currently tested in clinical studies \todo{ref. Maybe PMID:28733523 \cite{Espindola-Netto2017}?}. The major healthy tissue expressing NNMT is the liver, while most other tissues do not express NNMT or only at very low levels. Based on our analysis, we would furthermore suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors. This is furthermore relevant, as the missing degradation in those cells would potentially lead to an accumulating of Nam that could outcompete the inhibitor in those cells. But these aspects require further investigation.

Our combined phylogenetic/modelling analysis furthermore provides a potential explanation both for the co-occurrence of NADA and NamPRT in bacteria, but even more strongly for the loss of NADA in vertebrates. We here show that when simulating compartments that share the same limited source of Nam, the compartment that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate as the compartment containing NADA. The combination of a high affinity NamPRT with NNMT, which is the dominant enzyme combination found in vertebrates, seems to provide a competitive advantage. As this may also hold for mammalian-associated bacteria, particularly pathogens, we wanted to see whether pathogenic bacteria solely express NamPRT. Unfortunately, bacterial habitat information is currently far from complete and often difficult to access especially at the NCBI database that was used for the analysis provided here, as it is the most comprehensive sequence database to date. We therefore manually looked through the bacteria found to encode for NamPRT and indeed found that many but not all of them have been characterised to be pathogenic. It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree\todo{ref. Maybe PMID:19721089 \cite{Gazzaniga2009}?}.

>>>>

Our findings shed also new light on the potential physiological role of NNMT, which has earlier been recognized as potential marker for some types of cancer (e.g. \cite{Okamura1998}). It might well be that increased NNMT expression in cancer might serve to remove Nam produced by increased NAD-dependent signalling. To maintain high NAD concentrations a simultaneous overexpression of NamPRT is required, which is what has been found in some types of cancer \cite{Bi2011,Wang2011}. As NNMT does, however, provide an advantage as long as NamPRT affinity is sufficiently high, this might provide an useful avenue for treatment, especially for types of cancer that express NNMT at a high level. These would be more susceptible to competitive inhibitors of NamPRT, and several of those inhibitors are currently tested in clinical studies \cite{Espindola-Netto2017} \todo{Are there other refs?}. The major healthy tissue expressing NNMT is the liver, while most other tissues do not express NNMT or only at very low levels. Based on our analysis, we would furthermore suggest that it might be reasonable to screen patients before treatment, as non-NNMT expressing tumours might respond less to competitive NamPRT inhibitors. This is furthermore relevant, as the missing degradation in those cells would potentially lead to an accumulating of Nam that could outcompete the inhibitor in those cells. But these aspects require further investigation.

Our combined phylogenetic/modelling analysis furthermore provides a potential explanation both for the co-occurrence of NADA and NamPRT in bacteria, but even more strongly for the loss of NADA in vertebrates. We here show that when simulating compartments that share the same limited source of Nam, the compartment that contains NamPRT and NNMT has a higher steady state NAD concentration and NAD consumption rate as the compartment containing NADA. The combination of a high affinity NamPRT with NNMT, which is the dominant enzyme combination found in vertebrates, seems to provide a competitive advantage. As this may also hold for mammalian-associated bacteria, particularly pathogens, we wanted to see whether pathogenic bacteria solely express NamPRT. Unfortunately, bacterial habitat information is currently far from complete and often difficult to access especially at the NCBI database that was used for the analysis provided here, as it is the most comprehensive sequence database to date. We therefore manually looked through the bacteria found to encode for NamPRT and indeed found that many but not all of them have been characterised to be pathogenic. It should be noted that the distribution of NADA and NamPRT does not follow the bacterial species tree \cite{Gazzaniga2009}.

### 2018-07-10 latex/figurelegends.tex
Resolved many todos and typos in discussion.
Incorporated changes suggested by Dorothée

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{fig_6_unresolved_loop}
  \caption{\textbf{The function of the structurally unresolved loop structure of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A). Coloured circles indicate the enzymes present in the respective species blue: NamPRT and NNMT; black: NamPRT, NADA and NNMT; yellow: NamPRT and NADA. For a more comprehensive alignment, please see supplementary figure~S1. The visualisation of human NamPRT (B) is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} using the model 2H3D of the human NamPRT as template \cite{Wang2006}. The inserted region is not resolved in any of currently available crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red. Immunofluorescence images (C) show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop  is not changed compared to FLAG-tagged human wildtype NamPRT. Both show a heterogeneous nuclear-cytosolic localisation in HeLa S3 cells. \textit{In vitro} measurements using recombinant protein show that the mutant NamPRT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E). Bars in panels D and E that have different letters indicate significant difference of measured values as estimated using a T test assuming independent samples and significance at $p < 0.05$.}
  \label{fig:unresolved_loop}
\end{figure}

@@

\todo{What about the positive selection in NNMT?}

>>>>

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.7\textwidth,draft]{fig_6_unresolved_loop}
  \caption{\textbf{The function of the structurally unresolved loop of NamPRT.} Most Deuterostomia that encode NNMT show a sequence insertion in the N-terminal region of NamPRT that has been revealed by multiple sequence alignment of NamPRT from different species (A). Coloured circles indicate the enzymes present in the respective species blue: NamPRT and NNMT; black: NamPRT, NADA and NNMT; yellow: NamPRT and NADA. For a more comprehensive alignment, please see supplementary figure~S1. The visualisation of human NamPRT (B) is based on a structure prediction of SWISS-MODEL \cite{Arnold2006,Biasini2014} using the model 2H3D of the human NamPRT as template \cite{Wang2006}. The inserted region is not resolved in any of currently available crystal structures of NamPRT and thus appears to be a flexible loop structure at the surface of the NamPRT dimer, coloured in red. Immunofluorescence images (C) show that the localisation of the FLAG-tagged mutant protein lacking the unresolved loop  is not changed compared to FLAG-tagged human wildtype NamPRT. Both show a heterogeneous nuclear-cytosolic localisation in HeLa S3 cells. \textit{In vitro} measurements using recombinant protein show that the mutant NamPRT has a lower activity than the wildtype enzyme (D) and is not activated by ATP (E). Bars in panels D and E that have different letters indicate significant difference of measured values as estimated using a T test assuming independent samples and significance at $p < 0.05$.}
  \label{fig:unresolved_loop}
\end{figure}

@@

### 2018-07-10 latex/header.tex
Incorporated changes suggested by Dorothée

\noindent
\textsuperscript{2}Department of Molecular Biology, University of Bergen, Thormøhlensgt. 55, 5020 Bergen, Norway

@@

NAD is best known as cofactor in redox reactions, but it is also substrate of NAD-dependent signalling reactions that consume NAD and release nicotinamide (Nam). In eukaryotes, two different Nam salvage pathways exist. Phylogenetic analysis of the pathway show that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates. Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).  Despite the occurrence of this Nam degrading enzyme, we observe a diversification of NAD dependent signalling enzymes in vertebrates. This seems counterintuitive.

>>>>

\noindent
\textsuperscript{2}Department of Biological Sciences, University of Bergen, Thormøhlensgt. 55, 5020 Bergen, Norway

@@

NAD is best known as cofactor in redox reactions, but it is also substrate of NAD-dependent signalling reactions that consume NAD and release nicotinamide (Nam). In eukaryotes, two different Nam salvage pathways exist. Phylogenetic analysis of the pathway shows that in lower organisms the initial deamidation of Nam is prevalent, whilst the direct conversion of Nam to the mononucleotide by Nam phosphoribosyltransferase (NamPRT) dominates in animals and eventually remains as the single Nam recycling route in vertebrates. Strikingly, loss of the deamidation pathway in early vertebrates is preceded by the occurrence of a new enzyme that marks Nam for excretion by methylation – nicotinamide N-methyltransferase (NNMT).  Despite the occurrence of this Nam degrading enzyme, we observe a diversification of NAD dependent signalling enzymes in vertebrates. This seems counterintuitive.

### 2018-07-10 latex/methods.tex
Incorporated changes suggested by Dorothée

\section{Experimental Procedure}

@@

For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites. Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 (NamPRT-$\Delta$42-51). For prokaryotic expression with an N-terminal 6xHis-tag, the wild-type and mutant ORFs were inserted into pQE-30 (Qiagen) via BamHI and PstI-sites. All cloned sequences were verified by DNA sequence analysis.

@@

The spectra phase and baseline were automatically and manually corrected using TopSpin 3.5 software (Bruker Biospin). Quantification of nicotinamide mononucleotide (NMN) was done by the integration of the peak at 9.57\,ppm and DSS used as an internal standard.

>>>>

\section{Experimental Procedures}

@@

For eukaryotic expression with a C-terminal FLAG-epitope, the open reading frame (ORF) encoding human NamPRT was inserted into pFLAG-CMV-5a (Merck - Sigma Aldrich) via EcoRI/BamHI sites. Using a PCR-based approach, this vector provided the basis for the generation of a plasmid encoding a NamPRT deletion mutant lacking amino acid residues 42-51 ($\Delta$42-51 NamPRT). For prokaryotic expression with an N-terminal 6xHis-tag, the wild-type and mutant ORFs were inserted into pQE-30 (Qiagen) via BamHI and PstI-sites. All cloned sequences were verified by DNA sequence analysis.

@@

The spectra phase and baseline were automatically and manually corrected using TopSpin 3.5 software (Bruker Biospin). Quantification of nicotinamide mononucleotide (NMN) was done by the integration of the peak at 9.52\,ppm and DSS used as an internal standard.

### 2018-07-10 latex/results.tex
Incorporated changes suggested by Dorothée

We have previously shown that NMNAT is essential for NAD biosynthesis independent of the biosynthetic precursor or pathway \cite{DeFigueiredo2011}. In addition, the Preiss-Handler pathway was found present in virtually all organisms included in an earlier phylogenetic analysis \cite{Gossmann2012FEBS}. Thus, we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT. The phylogenetic distribution of the two enzymes is scattered in bacteria \todo{MathiasB: please find and insert citation} and have been shown to occur together in some marine invertebrates \cite{Gossmann2012FEBS}. In addition, we included the Nam methyltransferase NNMT into our analysis, as its function and origin is unclear.

As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT. In contrast, Metazoa predominantly lost NADA amd have NamPRT together with NNMT. NNMT seems to have arisen \textit{de~novo} in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B). As we were unable to find any \textit{NNMT} gene with an e-value below 0.1 in fungi or plants. The gene with the highest sequence similarity in \textit{S.~cerevisiae} is ... with proven function ... \todo{Mathias B: include values} \todo{Answer from MB: There was not a single hit outside Metazoa}.

@@

The second hypothesis was that the sequence insertion influences the affinity of NamPRT, which would be in line with the predictions from our simulations. We recombinantly expressed both wildtype and mutant NamPRT in \textit{E.~coli}. After purification (suppl. fig.~S3), we measured the activity using NMR detection of NMN in the presence and absence of ATP. The enzymatic activity of the mutant enzyme lacking the loop was much lower than the activity of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D). Furthermore, an increase of the substrate concentration had a much stronger effect on the activity of the mutant compared to the wildtype enzyme, indicating a lower affinity. In contrast to wildtype NamPRT, the mutant did not exhibit an increased activity in the presence of ATP (fig.~\ref{fig:unresolved_loop}E).

>>>>

We have previously shown that NMNAT is essential for NAD biosynthesis independent of the biosynthetic precursor or pathway \cite{DeFigueiredo2011}. In addition, the Preiss-Handler pathway was found present in virtually all organisms included in an earlier phylogenetic analysis \cite{Gossmann2012FEBS}. Thus, we here focused on the phylogenetic analysis of the two enzymes that initiate the two different NAD salvage pathways, NADA and NamPRT. The phylogenetic distribution of the two enzymes is scattered in bacteria \cite{Gazzaniga2009} and have been shown to occur together in some marine invertebrates \cite{Gossmann2012FEBS}. In addition, we included the Nam methyltransferase NNMT into our analysis, as its function and origin is unclear.

As shown in Figure~\ref{fig:phylo_distribution}A, bacteria, fungi, and plants predominantly possess NADA and only a few of them harbour NamPRT. In contrast, Metazoa predominantly lost NADA amd have NamPRT together with NNMT. NNMT seems to have arisen \textit{de~novo} in the most recent common ancestor of Ecdysozoa and Lophotrochozoa (fig.~\ref{fig:phylo_distribution}B). As we were unable to find any \textit{NNMT} gene with an e-value below 0.1 in fungi or plants. The gene with the highest sequence similarity in \textit{S.~cerevisiae} is ... with proven function ... \todo{Mathias B: include values} \todo{Answer by MB: There was not a single hit outside Metazoa}.

@@

The second hypothesis was that the sequence insertion might influence NamPRT affinity. We thus expressed both wildtype and mutant proteins, N-terminally fused to a 6xHis-tag in \textit{E.~coli}, and purified them. The size exclusion chromatography profile showed that both wildtype and mutant protein were expressed as dimers (see suppl. fig.~S3), indicating that the missing residues in the mutant did not dramatically affect the protein structure. The enzymatic activity was measured by NMR spectroscopy using the detection of NMN produced in the presence and absence of ATP. Upon incubation with the NamPRT inhibitor FK866 \cite{Hasmann2003} for 30 minutes, both wildtype and mutant NamPRT did not produce any NMN, suggesting that binding of FK866 is not affected by the mutation (see suppl. fig.~S4). Nevertheless, the enzymatic activity of the mutant enzyme was approximately twice as low than that of the wildtype enzyme (fig.~\ref{fig:unresolved_loop}D). In contrast to wildtype NamPRT, the activity of the mutant enzyme was not stimulated in the presence of ATP (fig.~\ref{fig:unresolved_loop}E). Our observations suggest that the mutant enzyme is catalytically active, retains its dimeric state and sensitivity to FK866. However, the lower activity and the absence of catalytic activation by ATP indicate that deletion of amino acids 42 to 51 might have affected substrate binding.

### 2018-07-10 latex/supplementaries.tex
Incorporated changes suggested by Dorothée

>>>>

\subsection*{Supplementary figure S4}

\textbf{NMR measurement of NamPRT activity.} NamPRT activity was measured as described in Experimental Procedures and product (NMN) formation was detected using 1D 1H NMR spectroscopy. Inset on the right: molecular structure of NMN with the atom detected by NMR indicated by an arrow. The range used for NMN detection in typical 1D 1H NMR spectra of the enzymatic reactions is shown. NMN quantification was done with the singlet detected at 9.52 ppm. From the top to the bottom, peak detection of NMN standard (200\,µM), wildtype NamPRT (1\,mM Nam and 1\,mM PRPP), mutant $\Delta$42-51 NamPRT (1\,mM Nam and 1\,mM PRPP), wildtype NamPRT with FK866, and mutant $\Delta$42-51 NamPRT with FK866. Incubation with inhibitor FK866 was done for 30\,min at 30\,°C.
